A Simple, Useful, and Only Slightly Wrong Model of Virtual Memory

### 2010-10-06 article.tex
References and cleanup

The penalty for a page fault in virtual memory can be extreme, and can make
previously-small constants quite large. Thus, it would be of great use to be
able to easily estimate, for a given algorithm, the number of times the
algorithm will have to wait for the operating system to retrieve virtual memory
pages off the hard drive.  On a modern system, the CPU can perform 2 billion
operations a second, but the access time to retrieve a page of virtual memory
from a consumer hard drive can be multiple milliseconds.  Even if we assume
ultra-fast hard drives, a lower bound of 1 millisecond seems more than
resonable, which mean that the process is stalled, waiting for memory, for more
than 2 million wasted CPU operations.  If each tick of the CPU clock took a
second, then a virtual memory page fault would impose a penalty of more than 23
days!  Clearly, it would be good to be able to asymptotically compare the
behavior of different algorithms in the presence of virtual memory, so as to
minimize the number of page faults.

@@

>>>>

The penalty for a page fault in virtual memory can be extreme, and can make
previously-small constants quite large. Thus, it would be of great use to be
able to easily estimate, for a given algorithm, the number of times the
algorithm will have to wait for the operating system to retrieve virtual memory
pages off the hard drive.  On a modern system, the CPU can perform 2 billion
operations a second, but the access time to retrieve a page of virtual memory
from a consumer hard drive can be multiple milliseconds.  Even if we assume
ultra-fast hard drives, a lower bound of 1 millisecond seems more than
resonable, which mean that the process is stalled, waiting for memory, for more
than 2 million wasted CPU operations.  If each tick of the CPU clock took a
second, then a virtual memory page fault would impose a penalty of more than 23
days!  Clearly, it would be good to be able to asymptotically compare the
behavior of different algorithms in the presence of virtual memory, so as to
minimize the number of page faults --- either asymptotically or just as a leading constant\cite{queue}.

@@

\appendix

\section{Code availability}

All code written, all experimental data collected, and even this article may be found online at \url{http://github.com/pboothe/vmanalysis}
